---
title: "Microbome_SubsetASVs"
author: "Chloee McLaughlin"
date: "May 2022"
output: html_document
---
This script analyzes specific ASVs of Massila and Oxalobacteraceae. Relating microbiome variation to environment of origin.
```{r,load libraries, include=FALSE}
library(dplyr)
library(raster)
library(tidyr)
library(Boruta)
library(randomForest)
library(purrr)
library(hydroGOF)
library(rnaturalearth)
library(dismo)
library(sf)
library(gstat)
library(rex)
library(deldir)
library(readxl)
library(gplots)
library(RColorBrewer)
```

```{r, set wd, include=FALSE}
setwd("~/OneDrive - The Pennsylvania State University/Research/Microbiome_Peng/code/")
```

```{r, CIMMyT data , include=FALSE}
micro_env <- read.csv("../data/Microbiome_EnvVars_CleanedImputed.csv") #environmental data 
micro_lr <- micro_env %>% filter(Germplasm=="Landrace") ##subset for landraces

foam_env <- read.csv("../data/FOAM_Env_UsedForMIcrobiomeRFPredictions.csv") ##CIMMYT FOAM 
```

Oxalobacteraceae data 

```{r, oxalobacteraceae load and split data, include=FALSE}
oxalobacteraceae <- read.csv("../data/Oxalobacteraceae.csv")

oxa_wide <- oxalobacteraceae %>% pivot_wider(names_from = c(biome, pos, trt),
                             values_from = Oxalobacteraceae)

oxa_master <- left_join(oxa_wide, micro_lr, by="Genotype") 
```

Massilia data

```{r, massilia load and split data, include=FALSE}
massilia <- read.csv("../data/Massilia.csv")

mas_wide <- massilia %>% pivot_wider(names_from = c(biome, pos, trt),
                             values_from = Massilia)

mas_master <- left_join(mas_wide, micro_lr, by="Genotype") 
```

Subsetted ASVs of each compartment/type/treatment

```{r, ASV load and split data, include=FALSE}
asv_rhiz_CK <- read_xlsx("../data/BLUEs_Bacteria_ASVs_rhizosphere.xlsx", sheet = 1)
asv_rhiz_CK$biome <- "bacteria"
asv_rhiz_CK$pos <- "rhizosphere"
asv_rhiz_CK$trt <-"CK"

asv_rhiz_D <- read_xlsx("../data/BLUEs_Bacteria_ASVs_rhizosphere.xlsx", sheet = 2)
asv_rhiz_D$biome <- "bacteria"
asv_rhiz_D$pos <- "rhizosphere"
asv_rhiz_D$trt <-"D"

asv_rhiz_LN <- read_xlsx("../data/BLUEs_Bacteria_ASVs_rhizosphere.xlsx", sheet = 3)
asv_rhiz_LN$biome <- "bacteria"
asv_rhiz_LN$pos <- "rhizosphere"
asv_rhiz_LN$trt <-"LN"

asv_rhiz_LP <- read_xlsx("../data/BLUEs_Bacteria_ASVs_rhizosphere.xlsx", sheet = 4) 
asv_rhiz_LP$biome <- "bacteria"
asv_rhiz_LP$pos <- "rhizosphere"
asv_rhiz_LP$trt <-"LP"

asv_root_CK <- read_xlsx("../data/BLUEs_Bacteria_ASVs_root.xlsx", sheet = 1)
asv_root_CK$biome <- "bacteria"
asv_root_CK$pos <- "root"
asv_root_CK$trt <-"CK"

asv_root_D <- read_xlsx("../data/BLUEs_Bacteria_ASVs_root.xlsx", sheet = 2)
asv_root_D$biome <- "bacteria"
asv_root_D$pos <- "root"
asv_root_D$trt <-"D"

asv_root_LN <- read_xlsx("../data/BLUEs_Bacteria_ASVs_root.xlsx", sheet = 3)
asv_root_LN$biome <- "bacteria"
asv_root_LN$pos <- "root"
asv_root_LN$trt <-"LN"

asv_root_LP <- read_xlsx("../data/BLUEs_Bacteria_ASVs_root.xlsx", sheet = 4)
asv_root_LP$biome <- "bacteria"
asv_root_LP$pos <- "root"
asv_root_LP$trt <-"LP"

asv_all <- bind_rows(asv_rhiz_CK,asv_rhiz_D,asv_rhiz_LN,asv_rhiz_LP,asv_root_CK,asv_root_D,asv_root_LN,asv_root_LP)

asv_sub <- asv_all %>% dplyr::select(Genotype, biome, pos, trt, ASV10, ASV111, ASV1751, ASV18, ASV37, 
                              ASV49, ASV51, ASV56, ASV7, ASV79, ASV824)

asv_wide <- asv_sub %>% pivot_wider(names_from = c(biome,pos,trt),
                             values_from = c(ASV10, ASV111, ASV1751, ASV18, ASV37, 
                              ASV49, ASV51, ASV56, ASV7, ASV79, ASV824))
```

PCA for subsetted ASVs by treatment to see if increases signal as combined vs individual ASVs

```{r, Combined select column, echo=FALSE}
#asv_rhiz_CK
asv_rhiz_CK.pca <- asv_rhiz_CK %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$rhiz_CK.PC1 <-asv_rhiz_CK.pca[["x"]][,1]
asv_wide$rhiz_CK.PC2 <-asv_rhiz_CK.pca[["x"]][,2]

#asv_rhiz_D
asv_rhiz_D.pca <- asv_rhiz_D %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$rhiz_D.PC1 <-asv_rhiz_D.pca[["x"]][,1]
asv_wide$rhiz_D.PC2 <-asv_rhiz_D.pca[["x"]][,2]

##asv_rhiz_LN
asv_rhiz_LN.pca <- asv_rhiz_LN %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$rhiz_LN.PC1 <-asv_rhiz_LN.pca[["x"]][,1]
asv_wide$rhiz_LN.PC2 <-asv_rhiz_LN.pca[["x"]][,2]

##asv_rhiz_LP
asv_rhiz_LP.pca <- asv_rhiz_LP %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$rhiz_LP.PC1 <-asv_rhiz_LP.pca[["x"]][,1]
asv_wide$rhiz_LP.PC2 <-asv_rhiz_LP.pca[["x"]][,2]

#asv_root_CK
asv_root_CK.pca <- asv_root_CK %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$root_CK.PC1 <-asv_root_CK.pca[["x"]][,1]
asv_wide$root_CK.PC2 <-asv_root_CK.pca[["x"]][,2]

#asv_root_D
asv_root_D.pca <- asv_root_D %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$root_D.PC1 <-asv_root_D.pca[["x"]][,1]
asv_wide$root_D.PC2 <-asv_root_D.pca[["x"]][,2]

#asv_root_LP
asv_root_LP.pca <- asv_root_LP %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$root_LP.PC1 <-asv_root_LP.pca[["x"]][,1]
asv_wide$root_LP.PC2 <-asv_root_LP.pca[["x"]][,2]

#asv_root_LN
asv_root_LN.pca <- asv_root_LN %>% 
  dplyr::select(Genotype, ASV10, ASV111, ASV1751, ASV18, ASV37, ASV49, ASV51, ASV56, ASV7, ASV79, ASV824) %>% 
  dplyr::select(where(is.numeric)) %>% 
  prcomp(scale=TRUE)

asv_wide$root_LN.PC1 <-asv_root_LN.pca[["x"]][,1]
asv_wide$root_LN.PC2 <-asv_root_LN.pca[["x"]][,2]
```

Combine phenotypic/env data

```{r, ASV df, echo=FALSE}
asv_master <- left_join(asv_wide, micro_lr, by="Genotype")
```

FEATURE SELECTION - Identify environmental descriptors that describe variation in all massilia and oxalobacteraceae ASV abundance
The results for these feature selection sections can also be loaded by reading in the RDS objects on lines 253-255

```{r, feature selection, echo=FALSE}
micro_bind <- as.data.frame(na.omit(mas_master)) #switch micro_bind to oxa_master to get oxalobacteraceae values
#micro_bind <- as.data.frame(na.omit(oxa_master)) 
set.seed(111)
y=NULL

for (i in (2:9)) {
    print(names(micro_bind)[i]) #Print name of the env var to be tested 
    boruta_train <- Boruta(micro_bind[,i] ~., data = micro_bind[,(c(i,16:169))], maxRuns=100,doTrace = 0) 
    boruta_signif <- getSelectedAttributes(boruta_train, withTentative = FALSE) #Get only the important vars
    tmp <- toString(boruta_signif) #Change significant vars to a string 
    print(boruta_signif) #Print sig env variables
    y <- rbind(y, tmp) #make a matrix of important variables 
    }

micro_boruta <- as.matrix(y) #Rename matrix
colnames(micro_boruta) <- c("Vars") #Name variables column (stored as a string)
micro_boruta <- subset(micro_boruta, grepl(rex(at_least(group(",", anything), 2)), micro_boruta[,1])) #Only include traits w/ more than 2 aspects of environment describing variation
rownames(micro_boruta) <- gsub(",.*$", "", as.matrix(micro_boruta)) #make variables row names

mas_boruta <- micro_boruta
#oxa_boruta <- micro_boruta
```

FEATURE SELCTION - Subsetted ASVs of importance individually (ASV#) and combined (PCs of focal ASVs)

```{r, feature selection, echo=FALSE}
micro_bind <- as.data.frame(na.omit(asv_master))
set.seed(111)
y=NULL

for (i in (2:105)) {
    print(names(micro_bind)[i]) #Print name of the env var to be tested 
    boruta_train <- Boruta(micro_bind[,i] ~., data = micro_bind[,(c(i,112:265))], maxRuns=100,doTrace = 0) 
    boruta_signif <- getSelectedAttributes(boruta_train, withTentative = FALSE) #Get only the important vars
    tmp <- toString(boruta_signif) #Change significant vars to a string 
    print(boruta_signif) #Print sig env variables
    y <- rbind(y, tmp) #make a matrix of important variables 
    }

micro_boruta <- as.matrix(y) #Rename matrix
colnames(micro_boruta) <- c("Vars") #Name variables column (stored as a string)
micro_boruta <- subset(micro_boruta, grepl(rex(at_least(group(",", anything), 2)), micro_boruta[,1])) #Only include traits w/ more than 2 aspects of environment describing variation
rownames(micro_boruta) <- gsub(",.*$", "", as.matrix(micro_boruta)) #make variables row names

asv_boruta <- micro_boruta
```

LOAD BORUTA OBJECTS MADE IN PREVIOUS SECTION - Shows the select ASVs, massilia, and oxalobacteraceae under specific treatments that have relationship to environmental descriptors 

```{r, feature selection loade, echo=FALSE}
#saveRDS(asv_boruta, "../output/Boruta_ASVsSUBSET.rds")
#saveRDS(mas_boruta, "../output/Bortua_massilia.rds")
#saveRDS(oxa_boruta, "../output/Boruta_oxalobacteraceae.rds")

asv_boruta <- readRDS("../output/Boruta_ASVsSUBSET.rds")
oxa_boruta <- readRDS("../output/Boruta_oxalobacteraceae.rds")
mas_boruta <- readRDS("../output/Boruta_massilia.rds")
```

RANDOM FOREST -  Build models, for saving space and automation purposes, I change out "imp" and "training_df" for whichever data frame of interest. All other other items in this section stay constant 

```{r, build models, make predicitons, echo=false}
##Focal ASVs
imp = asv_boruta ##feature selection matrix
training_df = na.omit(asv_master)  ##training/testing df 

##Massilia
#imp = mas_boruta ##feature selection matrix
#training_df = na.omit(mas_master) ##training/testing df 

##Oxalobacteraceae
#imp = oxa_boruta ##feature selection matrix
#training_df = na.omit(oxa_master )   ##training/testing df 

prediction_df = foam_env ##predicted df
x = NULL ##will be filled
output_table = NULL ##will be filled

#pdf("../output/ModelOutputs_ASVsSUBSET.pdf") #i in c(1:32,34:43,45,47:length(imp)) #have to skip columns where values stay close to constant (vary by <.0001), makes regression step difficult
#pdf("../output/ModelOutputs_oxalobacteraceae.pdf")
#pdf("../output/ModelOutputs_massilia.pdf")
#par(mar=c(2,2,2,2))
#par(oma=c(0,0,0,0))
#par(mfrow=c(4,4))

for (i in 1:length(imp)) { 
  set.seed(111) #reproducability
  
  ##Set up the data 
  boruta_vars <- imp[i,] #store line of string with phene + important vars
  boruta_vars <- strsplit(x=boruta_vars, ", ") #split the string
  trait <- rownames(imp)[i,] #store the phenotype in trait
  phene <- as.data.frame(training_df[,trait]) 
  rf.data <- training_df[,(boruta_vars[[1]])] #subset matching training data 
  names(rf.data)[1] <- "target" #make RF target
  
  ##Splitting data into training and testing  
  data_set_size = floor(nrow(rf.data)*0.70)
  index <- sample(1:nrow(rf.data), size=data_set_size)
  training <- rf.data[index,]
  testing <- rf.data[-index,]
  
  ##Run the model
  rf <- randomForest(target~., data=training, proximity=TRUE) 
  
  ##Scatter plots
  result <- data.frame(testing$target, predict(rf, testing[-1], type= "response"))
  xaxis = result[,1]
  yaxis = result[,2]
  all = c(xaxis,yaxis)
  range = c(min(all), max(all))
  plot(result, main=trait, cex.main=.75, xlab="Testing", ylab="RF Predicted", xlim=range,
       ylim=range)
  line <- lm(result[,2] ~ result[,1])
  abline(line, col="blue")
  legend("bottomright", bty="n", legend=paste("R2 =", 
                                        format(summary(line)$r.squared, digits=4)))
  
  ##Importance plot
  varImpPlot(rf, main=trait, cex.main=.75)

  ##Automated importance table
  import <- as.data.frame(rf$importance)
  import <- cbind(trait, var=row.names(import), import)
  x <- rbind(x,import)
  
  ##making a data frame of predicted valued for prediction_df  
  rf.testing <- prediction_df[names(rf.data[,-c(1)])]
  prediction_store <- predict(rf, newdata=rf.testing)
  prediction_df$trait.pred <- prediction_store
  names(prediction_df)[names(prediction_df) == "trait.pred"] <- paste(trait, '_pred', sep='')
  
  #Make the outputs table of model info
  coef.determination <- summary(line)$r.squared
  nse <- NSE(result[,1], result[,2])
  mae <- mae(result[,1], result[,2])
  mse <- mse(result[,1], result[,2])
  output_table <- rbind(output_table, c(trait, coef.determination, nse, mae, mse))
  }
#dev.off()

output_table <- as.data.frame(output_table)
names(output_table)[1] <- "trait"
names(output_table)[2] <- "coef.determ"
names(output_table)[3] <- "nse"
names(output_table)[4] <- "mae"
names(output_table)[5] <- "mse"

#write.csv(output_table, "../output/MicrobimeRFModelOutputs_ASVsSUBSET.csv")

#microbime_predictions <- prediction_df[,c(1:2,5:6,166:length(prediction_df))]
#write.csv(microbime_predictions, "../output/FOAMpredictions_ASVsSUBSET.csv")
#write.csv(microbime_predictions, "../output/FOAMpredictions_massilia.csv")
#write.csv(microbime_predictions, "../output/FOAMpredictions_oxalobacteraceae.csv")
```

Heatmap of model R-squared values

```{r, model heatmap, make predicitons, echo=false}
imp.table.micro <-read_excel("../output/MicrobimeRFModelOutputs_ASVsSUBSET.xlsx",sheet=2)

imp.table.micro <- as.data.frame(imp.table.micro)
row.names(imp.table.micro) <- imp.table.micro$trait
micro.imp <- as.matrix(imp.table.micro[,-c(1)])

my_palette <- colorRampPalette(brewer.pal(9, "YlOrRd"))(24)
coul <- colorRampPalette(brewer.pal(6, "YlOrRd"))(19)
mybreaks <- seq(0, .6, length.out=20)

heatmap.2(micro.imp, cexRow = 1, cexCol = .9, Colv = NA, Rowv = NA, margins =c(7,7), col=coul, breaks=mybreaks, tracecol=NA)
```
